---
title: "Analysis"
author: "Amber Thomas"
date: "2/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Is there any correlation between a student making it onto the top 100 hs players list and doing well in the NBA? Do they make it into the NBA more often? Can we predict their success? 

```{r}
library(tidyverse)
library(here)
```


## Importing Data

```{r}
draft <- read.csv("../output/draft.csv", header = TRUE, stringsAsFactors = FALSE)
overview <- read.csv("../output/overview.csv", header = TRUE, stringsAsFactors = FALSE)
allPlayers <- read.csv("../output/players.csv", header = TRUE, stringsAsFactors = FALSE)
seasons <- read.csv("../output/seasons.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Exploration

### What % of students on the Top 100 List in HS make it to the NBA? 

First, we can only keep players that were recruited before 2013

```{r}
players <- allPlayers %>% 
  filter((draft_year <= 2014 | is.na(draft_year)) & (recruit_year <= 2013 | is.na(recruit_year)))
```

```{r}
top100 <- players %>% 
  filter(!is.na(recruit_year))

toNBA <- players %>% 
  filter(!is.na(draft_year)) %>% 
  filter(!is.na(recruit_year)) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count / nrow(top100))

toNBA$percent
```

Alright, so about 27% of students from the Top 100 list make it to the NBA at all. 

What about students not in the Top 100 that make it to the NBA?

```{r}
notTop <- players %>% 
  filter(is.na(recruit_year))

lowToNBA <- players %>% 
  filter(!is.na(draft_year)) %>% 
  filter(is.na(recruit_year)) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count / nrow(notTop))

lowToNBA$percent
```
How many of the top 100 make it past two years? 

```{r}
top100 %>% 
  filter(valid_seasons >= 2) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count/toNBA$count)
```
12% of players from the Top 100 list play more than 2 seasons. 

And not in top 100 that make it past 2 years? 

```{r}
notTop %>% 
  filter(valid_seasons >= 2) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count/nrow(notTop))
```


### How many students not on the Top 100 list make it to the NBA? 

```{r}
notOnList <- players %>% 
  filter(is.na(recruit_year)) %>% 
  filter(valid_seasons >= 2)

contract <- draft %>% 
  filter(link != "")
  

# what % of people drafted, that played more at least 1 season were not on Top 100 for high school?
nrow(notOnList) / nrow(players)
```


How many were drafted and didn't get a contract? 

```{r}
noContract <- allPlayers %>% 
  filter(!is.na(draft_year)) %>% 
  filter(link == "")
```

What's the vorp distribution for each hs rank? 

```{r}
vorp <- top100 %>% 
  filter(valid_seasons >= 2) %>% 
  mutate(groupRank = rank - (rank %% 10))

ggplot(vorp, aes(x = groupRank, y = nba_median_vorp_rank, group = groupRank)) + geom_boxplot()
```

```{r}
ggplot(vorp, aes(x = groupRank, y = nba_median_pipm_rank, group = groupRank)) + geom_boxplot()
```

What about average rank for groups of vorp values? 

```{r}
vorpGroups <- top100 %>% 
  filter(valid_seasons >= 2) %>% 
  mutate(groupVorp = nba_median_vorp_rank - (nba_median_vorp_rank %% 25)) %>% 
  group_by(groupVorp) %>% 
  summarise(avgRank = median(rank),
            sd = sd(rank))
```

