---
title: "Analysis"
author: "Amber Thomas"
date: "2/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Is there any correlation between a student making it onto the top 100 hs players list and doing well in the NBA? Do they make it into the NBA more often? Can we predict their success? 

```{r}
library(tidyverse)
library(here)
```


## Importing Data

```{r}
draft <- read.csv("../output/draft.csv", header = TRUE, stringsAsFactors = FALSE)
overview <- read.csv("../output/overview.csv", header = TRUE, stringsAsFactors = FALSE)
allPlayers <- read.csv("../output/players.csv", header = TRUE, stringsAsFactors = FALSE)
seasons <- read.csv("../output/seasons.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Exploration

### What % of students on the Top 100 List in HS make it to the NBA? 

First, we can only keep players that were recruited before 2013

```{r}
players <- allPlayers %>% 
  filter((draft_year <= 2014 | is.na(draft_year)) & (recruit_year <= 2013 | is.na(recruit_year)))
```

```{r}
top100 <- players %>% 
  filter(!is.na(recruit_year))

toNBA <- players %>% 
  filter(!is.na(draft_year)) %>% 
  filter(!is.na(recruit_year)) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count / nrow(top100))

toNBA$percent
```

Alright, so about 27% of students from the Top 100 list make it to the NBA at all. 

What about students not in the Top 100 that make it to the NBA?

```{r}
notTop <- players %>% 
  filter(is.na(recruit_year))

lowToNBA <- players %>% 
  filter(!is.na(draft_year)) %>% 
  filter(is.na(recruit_year)) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count / nrow(notTop))

lowToNBA$percent
```
How many of the top 100 make it past two years? 

```{r}
top100 %>% 
  filter(total_seasons >= 2) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count/toNBA$count)
```
12% of players from the Top 100 list play more than 2 seasons. 

And not in top 100 that make it past 2 years? 

```{r}
notTop %>% 
  filter(total_seasons >= 2) %>% 
  summarise(count = n()) %>% 
  mutate(percent = count/nrow(notTop))
```


### How many students not on the Top 100 list make it to the NBA? 

```{r}
notOnList <- players %>% 
  filter(is.na(recruit_year)) %>% 
  filter(total_seasons >= 2)

contract <- draft %>% 
  filter(link != "")
  

# what % of people drafted, that played more at least 1 season were not on Top 100 for high school?
nrow(notOnList) / nrow(players)
```


How many were drafted and didn't get a contract? 

```{r}
noContract <- allPlayers %>% 
  filter(!is.na(draft_year)) %>% 
  filter(link == "")
```

What's the vorp distribution for each hs rank? 

```{r}
vorp <- top100 %>% 
  filter(total_seasons >= 2) %>% 
  mutate(groupRank = rank - (rank %% 10))

ggplot(vorp, aes(x = groupRank, y = nba_median_vorp_rank, group = groupRank)) + geom_boxplot()
```

```{r}
ggplot(vorp, aes(x = groupRank, y = nba_median_pipm_rank, group = groupRank)) + geom_boxplot()
```

What about average rank for groups of vorp values? 

```{r}
vorpGroups <- top100 %>% 
  filter(total_seasons >= 2) %>% 
  mutate(groupVorp = nba_median_vorp_rank - (nba_median_vorp_rank %% 25)) %>% 
  group_by(groupVorp) %>% 
  summarise(avgRank = median(rank),
            sd = sd(rank))
```

### Creating data for graphic

For each player, I need to know if they

* Were on the Top 100 list in HS
* Skipped college, stopped during/after college
* Were drafted by the NBA
* Signed their NBA contract
* Made it past their rookie contract
* how well did they do once they got there? 

```{r}
colleges <- players %>% 
  filter(!is.na(rank)) %>% 
  count(college, sort = TRUE) %>% 
  filter(college != "",
         n >= 20)
```

```{r}
export <- players %>% 
  rowwise() %>% 
  mutate(avgRank = mean(c(top_mean_vorp_rank, top_mean_wa_rank))) %>% 
  ungroup() %>% 
  mutate(top = ifelse(is.na(recruit_year), 0, 1),
         smallMult = ifelse(college %in% colleges$college & !is.na(rank), 1, 0),
         highSchool = case_when(
           # did they play beyond hs?
           college != "" | link != "" | draft_year != "" ~ 1,
           TRUE ~ 0
         ),
         coll = case_when(
           # if they didn't make it past hs, they'll get an NA here
           highSchool == 0 ~ NA_real_,
           # if they skipped college, they'll get a 2 here
           draft_year == recruit_year ~ 2,
           # if they definitely played past hs and have a draft year, then they made it past college
           (highSchool == 1 & draft_year != "") | (highSchool == 1 & valid_seasons > 0)  ~ 1,
           # otherwise they stopped in college
           TRUE ~ 0
         ),
         draft = case_when(
           # if they didn't make it past college, they'll get an NA here
           is.na(coll) | coll == 0 ~ NA_real_,
           # if they were drafted, did they have any valid seasons? 
           coll >= 1 & total_seasons > 0 ~ 1,
           # otherwise, they didn't make it past being drafted
           TRUE ~ 0
         ),
         rookie = case_when(
           # if they didn't make it past the draft, they'll get an NA here
           is.na(draft) | draft == 0 ~ NA_real_,
           # did they make it past the first 2 years? 
           draft == 1 & total_seasons >= 2 ~ 1,
           # otherwise, they didn't make it past their rookie contract
           TRUE ~ 0
         ),
         success = case_when(
           # if they didn't make it past their rookie year, they'll get an NA here
           is.na(rookie) | rookie == 0 ~ NA_character_,
           # if they made it this far, how good are they (based on avg. vorp score?)
           # allstars
           avgRank <= 30 ~ "allstar",
           # great players
           between(avgRank, 30, 60) ~ "great",
           # good players
           between(avgRank, 60, 120) ~ "good",
           # bad players
           (rookie == 1 & is.na(avgRank)) | avgRank > 120  ~ "bad",
           TRUE ~ NA_character_
         )) %>% 
  select(c(name, college, link, smallMult, rank, draft_year, draft_rd, draft_pk, recruit_year, avgRank, top, highSchool, coll, draft, rookie, success, valid_seasons))  %>% 
  mutate(val = 1) %>% 
  spread(success, val, fill = NA_real_) %>% 
  select(-"<NA>") %>% 
  mutate(highest = case_when(
           !is.na(rank) & highSchool == 0 ~ "highSchool",
           coll == 0 ~ "college",
           draft == 0 ~ "draft",
           rookie == 0 ~ "rookie",
           bad > 0 ~ "bad",
           good > 0 ~ "good",
           great > 0 ~ "great",
           allstar > 0 ~ "allstar",
           TRUE ~ "other"
         ))

write.csv(export, here::here("processed_data", "playerPaths.csv"), row.names = FALSE, na = "")
```

```{r}
colleges <- players %>% 
  count(college, sort = TRUE) %>% 
  filter(college != "",
         n >= 20)
```

```{r}
check <- export %>% 
  filter(valid_seasons > 0 & is.na(draft_year))
```




